The interactions of rational, pragmatic agents  \n lead to efficient language structure and use

### 2019-01-29 paper/paper.Rmd
updates after most of MF / MWu comments

---
title: "Interactions of rational, pragmatic agents lead to efficient language structure and use"
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"

@@

The current work examines the question, “how does having \textit{lexical ambiguity} make using language $\ell$ more or less useful for communication”? We begin with a high-level introduction to the modelling framework introducing the basic ingredients we will need to represent language as repeated reference games. Following this introduction we derive a simple objective function for measuring the efficiency of linguistic systems in this setting. We move on to two case-studies exploring the questions posed above -- examining the efficiency of ambiguity from a language design and language use perspective.\par

@@

### Reference games
Zipf’s example of optimal speaker- and listener-languages took the form of a reference game (similar to those described by Wittgenstein). We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this setting, speakers and listeners are aware of a set of objects $M$, which we will refer to as *meanings* and are knowledgeable about the set of possible signals $U$ (*utterances*) that can be used to refer to a given meaning (see Figure 1 for an example reference game). Utterances may have different relative costs, operationalized via a prior over utterances $P(U)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $P(M)$\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work Regier, Kemp \& Kay (2015).}. We consider a set of contexts $C$ with an associated prior $P(C)$. Contexts describe different orderings of the underlying meaning prior, formalized as different conditional distributions over meanings e.g. $p(M|C=c_i)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m> = e$ is a tuple of utterance, meaning pairs.\par

### Languages
A language $\ell$ defines the set of semantic mappings between utterance and meanings. For example, figure 1 contains four utterances $U = \{\text{blue}, \text{green}, \text{square}, \text{circle}\} \text{ and three meanings }M = \{\text{green-square}, \text{blue-square}, \text{green-circle}\}$. The boolean matrix describes the literal semantics of the language $\ell$. Note that will consider language to "contain ambiguous items" if there is at least one utterance $u \in U$ which can apply to multiple meanings (i.e. $|[[u_i]]| > 1$). For example, in Figure 1 both the words "blue" and "green" are ambigous so we would say that $\ell$ contains ambiguity.

### Speakers and listeners
The Rational Speech-act framework (RSA) is computational-level theory of pragmatic language use. RSA can be understood as a formalization of essential Gricean pragmatic principles -- interlocutors reason about one another and their shared context. For this reason, we adopt RSA as our representational framework to operationalize Gricean (rational and pragmatic) speaker-listeners [@Grice1975a]. We refer the reader to Section 1 of our supplementary materials for a more detailed description of basic RSA speaker-listener definitions, and to Goodman & Frank [-@GoodmanFrank2016a] for an overview of its appliations. We limit the current discussion to essential, high-level details of the framework.\par

@@

Ambiguity is ubiquitous in natural language. The task of understanding language is marked by a frequent need to handle lexical ambiguity (words often have multiple meanings) as well as syntactic ambiguity (sentences often have multiple parses), in addition to many other instances of ambiguity [@WasowPerforsBeaver2005a]. Chomsky [-@Chomsky2002a] famously claimed that the presence of ambiguity in natural language provides evidence that language has not been optimized for communication, stating “If you want to make sure that we never misunderstand one another, for that purpose language is not well designed, because you have such properties as ambiguity.” Note that this analysis is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$ we described earlier and define in terms of equation 2 in section 2.2 for our supplementary materials.\par

@@

We provide a toy example of their argument here. Say we have two objects in the world $m_1$ and $m_2$, and two languages $\ell_1$ and $\ell_2$. In language $\ell_1$, the low-cost utterance $u_1$ can be used to refer to both $m1$ and $m_2$ (i.e. $[[u_1]]_{\ell_1} = \{m_1, m_2 \}$). With language $\ell_2$, however, the low cost utterance $u_1$ can only refer to $m_1$ and a higher-cost utterance $u_2$ can refer to $m_2$ (i.e. $[[u_1]]_{\ell_2} = \{m_1\}$ and $[[u_2]]_{\ell_2} = \{m_2 \}$). While it is cheaper for a speaker to use $\ell_1$ (because she can always use a low-cost utterance), it is more difficult for a listener (because $u_1$ is ambiguous). If context isn't disambiguating then we might prefer $\ell_2$ to $\ell_1$. But, if context is disambiguating then the speaker can use $u_1$ to refer to either $m_1$ or $m_2$ and $\ell_1$ should be preferred.\par

@@

The argument outlined by Piantados et al. [-@Piantadosi2011a] is compatible with our current reference game setting with an addition -- context. The authors argue that mapping two meanings $m_1$ and $m_2$ to a single utterance $u_1$ is useful when they can be disambiguated *in context*. In our case we consider a context $c_i \in C$ to specify a unique ordering of the need probabilities. That is, $p(m|c_1) \neq p(m|c_2)$, when there are two contexts $|C| = \{c_1, c_2\}$.\par

This in turn, leads to an update to our linguistic efficiency objective as we now would like to consider the average speaker-listener effort over all contextualized communicative events (section 3 of supplementary materials).\par

@@

We conduct $N=2000$ simulations. For each simulation we enumerate the set of *valid* languages in which $|U|=|M|=4$. Note that a language $\ell \in L$ is "valid" so long as each possible meaning in $m \in M$ can be referred to by at least one form $u \in U$ (every column of $\ell$ has some non-zero assignment) and each form maps to at least one meaning (every column has some non-zero assignment). For a given simulation the goal is to find the language $\ell^*$ which minimizes our cross-entropy objective and then check to see if that language contains ambiguity.\par

@@

For comparison, we also examine properties of optimal languages under two additional objectives. Zipf [-@Zipf1949a] proposed that the optimal speaker language $\ell_{speaker}^*$ would only optimize speaker effort and the optimal listener language $\ell_{listener}^*$ would only optimize listener effort. We define these objectives using the first and second half of equation 1 and with greater detail in the supplementary materials (Section 2.2).\par

@@

Piantadosi’s claim leads to the following set of hypotheses: (1) The probability that an optimal language contains ambiguity under our Cross-Entropy objective should increase with the number of contexts (as $|C|\rightarrow\inf$). (2) The optimal speaker language under our “Speaker-only” objective should always contain ambiguity (maximally re-using the lowest-cost utterance). (3) The optimal listener language should never contain ambiguous material.

## Results

## Summary

Piantadosi et al. [-@Piantadosi2011a] argued that it is useful to re-use low-cost linguistic forms for multiple meanings when they can be disambiguated in context. We explored the degree to which this this property emerged during communication between rational, pragmatic agents, showing that as contextual information increased, the optimal languages under our Cross-Entropy objective were more likely to have ambiguous lexical items. Importantly, we did not see this effect under alternative objectives which only emphasized speaker- or listener effort alone.\par

Our first experiment made a fairly direct test of the communicative function of ambiguity theory. Sampling meaning, utterance, and context priors, we examined properties of the language $\ell^*$ which minimized our objective. Results indicated that ambiguity is an efficient property when context is informative. This simulation, however, assumed that our speaker and listener agents could always disambiguate items context. That is, in our four conditions both agents had perfect knowledge of the relevant conditional distributions ($P(M|C_i)$).\par

This may be a strong assumption for describing much of day-to-day communication -- we seldom begin a conversation with perfect knowledge of the current context of a conversation. Under Piantadosi's framework, if context is not informative with respect to meaning at time position $i$ in a discourse, but is informative at position $i+k$, then the speaker should avoid using ambigous material at position $i$, only using it at a future point, $k$ steps later in the conversation. In this toy example we can consider “context” as analogous to a “topic” of conversation. Intuitvely, we can imagine a scenario in which a reader is beginning a newspaper article. While they may have some knowledge about the article’s topic (perhaps from the title), they may not have complete knowledge of its contents, including the persons or events involved. In this setting, using an ambiguous pronoun early (position $i$) in the article can lead to misunderstanding without sufficient context to support disambiguation. But if by position $i+k$ there is sufficient contextual information it may be efficient for the writer to use ambiguous material.\par

We consider such a scenario in simulation two -- analyzing efficient *use* of ambiguity over discourse in a reference game setting. Unlike our first simulation, we now assume a single language $\ell$, which contains both ambiguous and unambiguous utterances. The ambiguous utterances are less costly than the unambiguous utterances. Crucially, we do not assume that the listener knows the particular topic ($c_{current}$) of the conversation \textit{a priori}. Rather, we assume that the listener has knowledge of the set of possible topics $C = \{c_1, \dots, c_k\}$, but \textit{does not know which one is currently being used by the speaker}. Formally, this means the listener does not have access to the correct conditional distribution over meanings $P(M|C=c_{current})$ at the start of the discourse. In this setting, a speaker minimizing production effort would like to use the low-cost material as frequently as possible. However, this will incur a disambiguation cost to the listener if they are uncertain of the current context $c_{current}$ (topic of conversation).\par

Over the course of a discourse, $D$ the listener tries to infer both the current topic, $c_{current}$, as well as the particular meaning $m$ of a given utterance $u$. That is we consider a modified listener model $L(m, c|u,D;\ell)$ and speaker model $S(u|m,c,D;\ell)$. Note that both the speaker and listener can track the history of previous utterances $D$. Importantly, a listener can attempt to infer the current topic of conversation $c_{current}$ using the discourse history $D$. Section 3 of the supplementary materials provides more detail on the speaker and listener models.\par

```{r plot-optimal-use, fig.env = "figure*", fig.pos = "h", fig.width=6, fig.height=4, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Left facet (A) shows the empirical probability that our speaker used an ambiguous utterance as a function of discourse position. As the discourse proceeds, only our Full Pragmatic speaker is more likely to use an ambiguous term. Plot B tracks the Cross-Entropy objective under our three speaker models. The objective declines in all cases as the listener infers the correct topic distribution. However the pragmatic speaker arrives as the lowest value, do to reduction in both speaker and listener effort. Plot C tracks speaker effort across the three models. Only the Full Pragmatic speaker shows declines in effort across the discourse as they increasinlgy use ambiguous items as the listener becomes more sure of the topic of conversation."}
img <- png::readPNG("figs/discourse_grid_plot.png")
grid::grid.raster(img)
```

We conduct $N=600$ simulations, generating discourses of length $|D|=30$ utterances with three different speaker models ($n=200$ each).  We consider a single language $\ell$ with $|U|=6$ and $|M|=4$ specified by the boolean matrix:

\begin{center}
\begin{tabular}{ c | c c c c} 
& $m_1$ & $m_2$ & $m_3$ & $m_4$ \\
\hline
$u_1$ & 1 & 0 & 0 & 0 \\
$u_2$ & 0 & 1 & 0 & 0 \\
$u_3$ & 0 & 0 & 1 & 0 \\
$u_4$ & 0 & 0 & 0 & 1 \\
$u_5$ & 1 & 1 & 0 & 0 \\
$u_6$ & 0 & 0 & 1 & 1 \\
\end{tabular}
\end{center}

We assume that $p(u_5) = p(u_6) > p(u_{1})=\dots=p(u_{4})$. That is, the two ambiguous utterances ($u_5$ and $u_6$) are less costly than the non-ambiguous utterances.\par

## Speaker agents
We consider three types of speaker models. In our \textit{Full pragmatics} model, we consider a speaker agent who reasons about her listener and also has complete recall of the set of utterances in the discourse $D$. This speaker believes that the listener may not know the current topic $c_{current}$ at the start of the discrouse, but can infer it with enough data. In this way, contextual information shared by the speaker and listener may increase over the discourse (as the listeners estimate of the current topic ($P(C=c|D)$) comes in line with the actual topic known to the speaker). We compare two baselines models. The first, a \textit{Partial pragmatics} baseline describes a speaker who reasons about a listener, but assumes they have no access to discourse history. The second is a \textit{No pragmatics} speaker who does not consider a listener at all, but produces utterances according to the underlying langauge semantics ($\ell$) and topic probabilities ($p(M|C=c_{current}$). We provide detailed descriptions of each model in the supplementary materials Section 3.\par

## Hypotheses
Under Piantadosi's claim it is efficient to use low-cost, ambiguous material when context is disambiguating. In the current setting, this corresponds to speaker behavior in which ambiguouous forms are avoided early in the discourse, but prefered later in the discourse once the listener is confident of the topic. We should expect this to be reflected in our \textit{Full pragmatics} model. By contrast, our \textit{Partial pragmatics} speaker does not believe that there is sufficient contextual information for the listener to disambiguate her utterances. This should lead to total  avoidance of ambiguous material over the discourse. Finally, our \textit{No pragmatics} speaker should use low-cost material greedily since she does not consider a listener's need to disambiguate whatsoever.\par

## Results

Figure 2, plot A shows the empirical probability that a speaker uses an ambiguous utterance as a function of discourse position. Note that the \text{No pragmatic} baseline uses ambiguous utterances frequently and at a constant rate over the discourse, the \text{Partial pragmatic} baseline avoids ambiguous utterances entirely (reasonining that the utterances cannout be disambiguated), but the \text{Full pragmatic} model avoids ambiguous material at first, but employing them gradually increasingly as the discourse proceeds. Figure 2, plot B tracks the speaker-listener Cross-Entropy objective. Note that the objective decreases for all three models. This decrease in all three models is primarily driven by the listener updating his belief about the actual topic ($P(C=c_{current}|D)$). Of particular importance is the value of the objective at the end of the discourse (position $i=30$). The difference in the Cross-Entropy between the \text{Full Pragmatic} and \text{Partial Pragmatic} models at the end of the discourse is driven by the reduction in speaker costs described in Figure 2, plot C. That is, while speaker effort remains constant in both \textit{No pragmatic} and \textit{Partial pragmatic} baselines, effort declines for the \text{Full Pragmatic} model as they are increasingly able to use ambiguous material later in the discourse.\par

## Summary

These simulations explored the efficient *use* of ambiguity over a discourse, in an alternative instantiation of the argument outlined by Piantadosi et al. [-@Piantadosi2011a]. In the first set of simulations we varied the set of languages, examining the properties of languages which minimized our cross-entropy objective under different context conditions. In the current simulation we assumed that the listener did not know the current context (topic) \textit{a priori}, but could infer it from the discourse. We examined speaker production behavior, tracking use of low-cost, but ambiguous material over the discourse. We found that early in the discourse our \textit{Full pragmatic} speaker avoided using ambiguous material, only using the material later when the context was known to the listener. This is consistent with Piantaodi's claim -- it is useful to use low-cost material, but only when context is disambiguating.

@@

How do the competing pressures of speakers and listeners give rise to the distributional forms found in natural language? Zipf [-@Zipf1949a] proposed that the asymmetry between speaker and listener costs gives rise to a range of properties at the level of the lexicon. Subsequent functionalist approaches to language science have provided additional evidence that aspects of language structure and language use appear to be optimized for efficiency. In Zipf’s famous example of an optimal speaker and listener language he appears indicate that such optimization occurs in in-the-moment, local interactions between speakers and listeners. We explore this idea in our current project, following theoretical work by Horn [-@Horn1984a] who proposed the speaker-listener pressures identified by Zipf can also be interpreted in terms of the maxims governing rational conversation proposed by [@Grice1975a]. To so, we explored the interactions of rational pragmatic agents as a framework for understanding efficient language structure and use. We focused on an argument on the communicative function of ambiguity put forth by Piantadosi et al. [-@Piantadosi2011a]. We derive a novel speaker-listener cross-entropy objective for measuring the efficiency of a language in a reference game setting showing that optimal languages are more likely to contain ambiguous material when context is informative. Further we show that rational pragmatic agents will *use* ambiguous material efficiently, only when such use is supported by context. The intent is that this project provides a demonstration for the types of functionalist theories that can be explored within this framework. Among other language properties open to this type of analysis are work examining semantic typology [@RegierKempKay2015a], the emergence of compositionality [@SmithKirbyBrighton2003a] and convention formation, as well as efficient examples of language use including the efficient use of reduction and redundancy [@LevyJaeger2007a; @GenzelCharniak2002a].

>>>>

---
title: "The interactions of rational, pragmatic agents  \n lead to efficient language structure and use"
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"

@@

The current work examines the question, “how does having \textit{lexical ambiguity} make using language $\ell$ more or less useful for communication”? We begin with a high-level introduction to the modelling framework introducing the basic ingredients we will need to represent language as repeated reference games. We derive anobjective function for measuring the efficiency of linguistic systems in this setting. We move on to two case-studies exploring the questions posed above -- examining the efficiency of ambiguity from a language design and language use perspective. The contributions of this work is to show how the reference game framework, in combination with formal modles of communication, can be used to study questions at a range of different timescales, from language use in-the-moment to language change across historical time.\par

@@

### Reference games
Zipf’s example of optimal speaker- and listener-languages took the form of a reference game. We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this setting, speakers and listeners are aware of a set of objects $M$, which we will refer to as *meanings* and are knowledgeable about the set of possible signals $U$ (*utterances*) that can be used to refer to a given meaning (see Figure 1). Utterances may have different relative costs, operationalized via a prior over utterances $P(U)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $P(M)$\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work Regier, Kemp \& Kay (2015).}. We consider a set of contexts $C$ with an associated prior $P(C)$. Contexts describe different orderings of the underlying meaning prior, formalized as different conditional distributions over meanings e.g. $p(M|C=c)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m> = e$ is a tuple of utterance, meaning pairs.\par

### Languages
A language $\ell$ defines the set of semantic mappings between utterance and meanings. For example, Figure 1 contains four utterances $U = \{\text{blue}, \text{green}, \text{square}, \text{circle}\} \text{ and three meanings }M = \{\text{green-square}, \text{blue-square}, \text{green-circle}\}$. The boolean matrix describes the literal semantics of the language $\ell$. Note that will consider language to "contain ambiguous items" if there is at least one utterance $u \in U$ which can apply to multiple meanings (i.e. $|[[u_i]]| > 1$)\footnote{We use double brackets $[[\dots]]$ to represent denotation}. For example, in Figure 1 both the words "blue" and "green" are ambigous so we would say that $\ell$ contains ambiguity.

### Speakers and listeners
The Rational Speech-act framework (RSA) is computational-level theory of pragmatic language use. RSA can be understood as a formalization of essential Gricean pragmatic principles -- agents reason about one another and their shared context. For this reason, we adopt RSA as our representational framework to operationalize Gricean (rational and pragmatic) speaker-listeners [@Grice1975a]. We refer the reader to Section 1 of our supplementary materials for a more detailed description of basic RSA speaker-listener definitions, and to Goodman & Frank [-@GoodmanFrank2016a] for an overview of its appliations.\par

@@

The task of understanding language is marked by a frequent need to handle various forms of ambiguity; lexical, syntactic, among others [@WasowPerforsBeaver2005a]. The ubiquity of this property, however, has been argued to provide evidence that langauges are not been optimized for communication\footnote{Chomsky (2002) famously claimed "If you want to make sure that we never misunderstand one another, for that purpose language is not well designed, because you have such properties as ambiguity." Note that this analysis is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$ we described earlier and define in terms of equation 2 in SI 2.2 .} [@Chomsky2002a].

@@

We provide a toy example of their argument here. Say we have two objects in the world $m_1$ and $m_2$, and two languages $\ell_1$ and $\ell_2$. In language $\ell_1$, the low-cost utterance $u_1$ can be used to refer to both $m_1$ and $m_2$ (i.e. $[[u_1]]_{\ell_1} = \{m_1, m_2 \}$). With language $\ell_2$, however, the low cost utterance $u_1$ can only refer to $m_1$ and a higher-cost utterance $u_2$ can refer to $m_2$ (i.e. $[[u_1]]_{\ell_2} = \{m_1\}$ and $[[u_2]]_{\ell_2} = \{m_2 \}$). While it is cheaper for a speaker to use $\ell_1$ (because she can always use a low-cost utterance), it is more difficult for a listener (because $u_1$ is ambiguous). If context isn't disambiguating then we might prefer $\ell_2$ to $\ell_1$. But, if context is disambiguating then the speaker can use $u_1$ to refer to either $m_1$ or $m_2$ and $\ell_1$ should be preferred.\par

@@

We adpot the framework above, and (following Piantadosi et al. 2011), add \textit{context} ($C$). The authors argue that mapping two meanings $m_1$ and $m_2$ to a single utterance $u_1$ is useful when they can be disambiguated *in context*. In our case we consider a context $c_i \in C$ to specify a permutation of the meaning probabilities. That is, $p(m|c_1) \neq p(m|c_2)$, when there are two contexts $|C| = \{c_1, c_2\}$.This in turn, leads to an update to our linguistic efficiency objective as we now would like to consider the average speaker-listener effort over all contextualized communicative events (see section 3 SI).\par

We proceed by generating languages with different amounts of contextual support (varying the size of $|C|$). We search the space of languages, asessing if the ones which minimize our objective contain ambiguity. If context leads to more efficient commnication, then optimal languages should be more likley to be ambiguous as the amount of context increases.\par

@@

We conduct $N=2000$ simulations. For each simulation we enumerate the set of *valid* languages in which $|U|=|M|=4$ (recall that $U$ is our set of utterances and $M$ our set of meanings). Note that a language $\ell \in L$ is "valid" so long as each possible meaning in $m \in M$ can be referred to by at least one form $u \in U$ (every column of $\ell$ has some non-zero assignment) and each form maps to at least one meaning (every column has some non-zero assignment)\footnote{For the current set of simulations we consider languages in which there are four or five true values -- in a "four" language ambiguity is not possible, but in a "five" language there are ambiguous mappings.}. For a given simulation the goal is to find the language $\ell^*$ which minimizes our cross-entropy objective and then check to see if that language contains ambiguity.\par

@@

For comparison, we also examine properties of optimal languages under two additional objectives. Zipf [-@Zipf1949a] proposed that the optimal speaker language $\ell_{speaker}^*$ would only optimize speaker effort and the optimal listener language $\ell_{listener}^*$ would only optimize listener effort. We define these objectives using the first and second half of equation 1 (see SI section 2.2. for more detail).\par

## Results and Discussion

We explored the degree to which ambiguity improved efficiency during communication between rational, pragmatic agents. We showed that as contextual information increased, the optimal languages under our Cross-Entropy objective were more likely to have ambiguous lexical items. Importantly, we did not see this effect under alternative objectives which only emphasized speaker- or listener effort alone.\par

While our results indicate that ambiguity is an efficient property of contextualized language these simulations assumed that speaker-listener agents could always disambiguate items in context. That is, in our four conditions both agents had perfect knowledge of the relevant conditional distributions ($P(M|C)$). This assumption may be too strong for describing much of day-to-day communication -- we seldom begin a conversation with perfect knowledge of the current context of a conversation. We explore a case in which the listener has imperfect knowledge of context at the start of the conversation in Simulation 2.\par

We shift time-scales in Simulation 2 -- instead of examining the objective landscape of a space of languages, we now consider a single language that contains ambiguity and examine how that property is used over a discourse. If context is not informative with respect to meaning early in a discourse (position $i$), but is at later positions at position (position $i+k$), then the speaker should avoid using ambigous material early $i$, only using it $k$ steps later in the conversation. In this toy example we can consider “context” as analogous to a “topic” of conversation. Intuitively, we can imagine a scenario in which a reader is beginning a newspaper article. While they may have some knowledge about the article’s topic (perhaps from the title), they may not have complete knowledge of its contents, including the persons or events involved. In this setting, using an ambiguous pronoun early (position $i$) in the article can lead to misunderstanding without sufficient context to support disambiguation. But if by position $i+k$ there is sufficient contextual information it may be efficient for the writer to use ambiguous material.\par

We consider consider a single language $\ell$, which contains both ambiguous and unambiguous utterances. The ambiguous utterances are less costly than the unambiguous utterances. Crucially, we do not assume that the listener knows the particular topic ($c_{current}$) of the conversation \textit{a priori}. Rather, we assume that the listener has knowledge of the set of possible topics $C = \{c_1, \dots, c_k\}$, but \textit{does not know which one is currently being used by the speaker}. Formally, this means the listener does not have access to the correct conditional distribution over meanings $P(M|C=c_{current})$ at the start of the discourse. In this setting, a speaker minimizing production effort would like to use the low-cost material as frequently as possible. However, this will incur a disambiguation cost to the listener if they are uncertain of the current context $c_{current}$ (topic of conversation).\par

Over the course of a discourse, $D$ the listener tries to infer both the current topic, $c_{current}$, as well as the particular meaning $m$ of a given utterance $u$. That is, we consider updated listener and speakers ($L(m, c|u,D;\ell)$, $S(u|m,c,D;\ell)$). Note that both the speaker and listener can track the history of previous utterances $D$. Importantly, an agent can attempt to infer the current topic of conversation $c_{current}$ using the discourse history $D$ (see SI section 3 for more detail).\par

```{r plot-optimal-use, fig.env = "figure*", fig.pos = "h", fig.width=6, fig.height=2.5, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "(A) shows the empirical probability that our speaker used an ambiguous utterance as a function of discourse position. (B) shows speaker effort across the three models. (C) shows the Cross-Entropy objective under our three speaker models."}
img <- png::readPNG("figs/discourse_grid_plot.png")
grid::grid.raster(img)
```

We conduct $N=600$ simulations, generating discourses of length $|D|=30$ utterances with three different speaker models ($n=200$ each). We consider a single language $\ell$\footnote{See SI for the matrix notation of this langauge.} with $|U|=6$ and $|M|=4$ in which two of the utterances $u_5$ and $u_6$ are ambiguous and that $p(u_5) = p(u_6) > p(u_{1})=\dots=p(u_{4})$. That is, the two ambiguous utterances ($u_5$ and $u_6$) are less costly than the non-ambiguous utterances. (Note that use of this particular language is not essential -- the results are broadly generalizable languages that contain ambiguity.)\par

## Speaker agents
We consider three types of speaker models. In our \textit{Full pragmatics} model, we consider a speaker agent who reasons about her listener and also has complete recall of the set of utterances in the discourse $D$. This speaker believes that the listener may not know the current topic $c_{current}$ at the start of the discrouse, but can infer it with enough data. In this way, contextual information shared by the speaker and listener may increase over the discourse (as the listeners estimate of the current topic ($P(C=c|D)$) comes in line with the actual topic known to the speaker). We compare two baselines models. The first, a \textit{Partial pragmatics} baseline describes a speaker who reasons about a listener, but assumes they have no access to discourse history. The second is a \textit{No pragmatics} speaker who does not consider a listener at all, but produces utterances according to the underlying langauge semantics ($\ell$) and topic probabilities ($p(M|C=c_{current}$) (see SI section 3 for details).\par

@@

We assess whether it is efficient to use low-cost, ambiguous material when context is disambiguating. In the current setting, this corresponds to speaker behavior in which ambiguous forms are avoided early in the discourse, but prefered later in the discourse once the listener is confident of the topic. We should expect this to be reflected in our \textit{Full pragmatics} model. By contrast, our \textit{Partial pragmatics} speaker does not believe that there is sufficient contextual information for the listener to disambiguate her utterances. This should lead to total  avoidance of ambiguous material over the discourse. Finally, our \textit{No pragmatics} speaker should use low-cost material greedily since she does not consider a listener's need to disambiguate whatsoever.\par

## Results and Discussion

Figure 2, (A) shows the empirical probability that a speaker uses an ambiguous utterance as a function of discourse position. The \text{No pragmatic} baseline uses ambiguous utterances frequently and at a constant rate over the discourse and the \text{Partial pragmatic} baseline avoids ambiguous utterances entirely (reasoning that the utterances cannot be disambiguated). But, the \text{Full pragmatic} model avoids ambiguous material at first, but employing them gradually increasingly as the discourse proceeds. (B) shows the speaker-listener Cross-Entropy objective. Note that the objective decreases for all three models. This decrease in all three models is primarily driven by the listener updating his belief about the actual topic ($P(C=c_{current}|D)$). Of particular importance is the value of the objective at the end of the discourse (position $i=30$). The difference in the Cross-Entropy between the \text{Full Pragmatic} and \text{Partial Pragmatic} models at the end of the discourse is driven by the reduction in speaker costs described in (C). That is, while speaker effort remains constant in both \textit{No pragmatic} and \textit{Partial pragmatic} baselines, effort declines for the \text{Full Pragmatic} model as they are increasingly able to use ambiguous material later in the discourse.\par

In Simulation 2 we explored how ambiguity could be employed efficiently during discourse. We assumed that the listener did not know the current context (topic) \textit{a priori}, but could infer it from the discourse history. We examined speaker production behavior, tracking use of low-cost, but ambiguous material over the discourse. Early in the discourse the \textit{Full pragmatic} speaker avoided using ambiguous material, only using the material later when the context was known to the listener -- it is useful to use low-cost material, but only when context is disambiguating.

@@

How do the competing pressures of speakers and listeners give rise to the distributional forms found in natural language? Zipf [-@Zipf1949a] proposed that the asymmetry between speaker and listener costs gives rise to a range of properties at the level of the lexicon. We explored the interactions of rational pragmatic agents as a framework for understanding efficient language structure and use. We focused on an argument on the communicative function of ambiguity [@Piantadosi2011a]. We derive a novel speaker-listener cross-entropy objective for measuring the efficiency of a language in a reference game setting showing that optimal languages are more likely to contain ambiguous material when context is informative. Further we show that rational pragmatic agents will *use* ambiguous material efficiently, only when such use is supported by context.\par

The framework employed here -- the analysis of the dynamics of repeated reference games between rational pragmatic agents -- provides a tool for the analysis of functionalist theories, allowing linkage across timescales of analysis from conversation [@LevyJaeger2007a; @GenzelCharniak2002a] to typology [@RegierKempKay2015a]. In theory, this framework provides extensive opportunity to vary properties of the languages and agents, however we have focused on only a small subset here. For example, examining how increasing or decreasing \textit{rationality} (instantiated through the RSA $\alpha$ parameter) or \textit{pragmatics} (instantiated through the RSA recursion depth) impacts these results is a clear avenue for future work. An important limitation of the current work is the broad assumption that language can be described through such reference games. The degree to which properties of non-referential language can be captured in such a framework remains an important target for future work.

\vspace{1em} \fbox{\parbox[b][][c]{7.3cm}{\centering All experiments, data, and analysis code are available in the public repository for the project: (link will be available upon acceptance)}} \vspace{1em}
\noindent
